### 2008-12-18 talk/ecoop2009/benchmarks.tex
finish benchmark section, only the tables are missing

To measure how the JIT handles object-oriented features, we wrote a very
simple benchmark that involves attribute lookups and method calls.  We have an
\emph{accumulator} object, which has a field \lstinline{value} and a method
\lstinline{add}.  The method \lstinline{add} takes a parameter and adds it the
field \lstinline{value}.

XXX: update to the new version

Our benchmark accepts a paramter \lstinline{n}, create an \emph{accumulator},
and repeatedly calls \lstinline{add} on it, passing numbers from
\lstinline{n-1} to \lstinline{0}.

The JIT is able to completely remove the overhead of object creation,
attribute lookups and method calls, and the generated code results in a simple
loop doing additions in-place.

Table XXX and figure XXX show the time spent to run the benchmark with various
input arguments. Again, we can see that the jitted code is up to 500 times
faster than the interpreted one.

>>>>

To measure how the JIT handles object-oriented features, we wrote a very
simple benchmark that involves attribute lookups and polymorphic method calls:

\begin{lstlisting}
def main(n):
    if n < 0:
        n = -n
        obj = new(value, accumulate=count)
    else:
        obj = new(value, accumulate=add)
    obj.value = 0
    while n > 0:
        n = n - 1
        obj.accumulate(n)
    return obj.value

def count(x):
    this.value = this.value + 1

def add(x):
    this.value = this.value + x
\end{lstlisting}

The two \lstinline{new} operations create an object with exactly one field
\lstinline{value} and one method \lstinline{accumulate}, whose implementation
is found in the functions \lstinline{count} and \lstinline{add}, respectively.
When calling a method, the receiver is implicity passed and can be accessed
through the special name \lstinline{this}.

The computation \emph{per se} is trivial, as it calculates either $-n$ or
$1+2...+n-1$, depending on the sign of $n$. The interesting part is the
polymorphic call to \lstinline{accumulate} inside the loop, because the VM has
no way to know in advance which method to call (unless it does flow analysis,
which could be feasible in this case but not in general).  The equivalent C\#
code we wrote uses two classes and a \lstinline{virtual} method call to
implement this behaviour.

However, our generated JIT does not compile the whole function at
once. Instead, it compiles and executes code chunk by chunk, waiting until it
knows enough informations to generate highly efficient code.  In particualr,
at the time when it emits the code for the inner loop it exactly knows the
type of \lstinline{obj}, thus it can remove the overhead of dynamic dispatch
and inline the method call.  Moreover, since \lstinline{obj} never escapes the
function, it is \emph{virtualized} and its field \lstinline{value} is stored
as a local variable.  As a result, the generated code results in a simple loop
doing additions in-place.

Table XXX show the time spent to run the benchmark with various input
arguments. Again, we can see that the jitted code is up to 500 times faster
than the interpreted one.  Moreover, the code generated by the JIT is
\textbf{faster} than the equivalent C\# code.

Probably, the C\# code is slower because:

\begin{itemize}
\item The object is still allocated on the heap, and thus there is an extra
  level of indirection to access the \lstinline{value} field.
\item The method call is optimized through a \emph{polymorphic inline cache}
  (XXX: citation needed), that requires a guard check at each iteration.
\end{itemize}

\anto{maybe we should move the following paragraph to
  abstract/introduction/conclusion?}

Despite being only a microbenchark, this result is very important as it proves
that our strategy of intermixing compile time and runtime can yield to better
performances than current techniques.  The result is even more impressive if
we consider dynamically typed languages as TLC are usually considered much
slower than the statically typed ones.
